---
title: "Assignment 3 - Our own data analysis"
author: "Akaran Sivakumar (AS), Lelia Rønnow (LR), Shiraz Ben-Shoshan (SBS)"
date: "01-12-2024"
output:
  html_document: default
  pdf_document: default
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages and data 
```{r}
library(tidyverse)
#install.packages("countrycode")
library(countrycode)
```


```{r}
# Setting working directory
#setwd("~/Desktop/AarhusUni/Semester5/IntroCultDataSci")

# Loading Data
WHO_MHExp_and_Deaths <- read_csv("archive/WHO_MHExp_and_Deaths.csv")
```
# Preliminary data visualisations (AS)
```{r}
WHO_MHExp_and_Deaths %>%
ggplot(aes(x = HExp_Pctage_Y,
y = Suicide_p100)) +
geom_bar(stat = "summary", fun = "mean") +
theme_minimal() +
labs(x = "Health Expenditure", y = "Suicide rate") +
ggtitle("Suicide rate and Health expenditure")
```
The plot above does not show any clear patterns, so we will now conduct further exploratory plots to discover hidden patterns.

```{r}
WHO_MHExp_and_Deaths %>%
ggplot(aes(x = MHExp_Pctage_2011,
y = Dep_Num_2015)) +
geom_point()+
theme_minimal() +
labs(x = "Mental Health expenditure index in 2011", y = "Depression estimate index in 2015") +
ggtitle("Mental heatlh index and Depression estimate index")
```
## Data mutation and further visualisations (LR)
```{r}
# Load the package
library(reshape2)

# Converting nominal deaths to a ratio of per 100k
WHO_MHExp_and_Deaths$Deaths_per_100k <- (WHO_MHExp_and_Deaths$Deaths_All_Types / WHO_MHExp_and_Deaths$Population) * 100000

# Convert country names to continent names
WHO_MHExp_and_Deaths$continents <- countrycode(WHO_MHExp_and_Deaths$Country_Name, origin = "country.name", destination = "continent")

# Remove the first and last columns
df_filtered <- WHO_MHExp_and_Deaths[, -c(1, ncol(WHO_MHExp_and_Deaths)-1)]

numeric_df <- df_filtered[, sapply(df_filtered, is.numeric)]
# Calculate the correlation matrix and convert it to a long format
cor_matrix <- cor(numeric_df, use = "complete.obs")
cor_long <- melt(cor_matrix)

# Correlation matrix plot
ggplot(cor_long, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1)) +
  theme_minimal() +
  labs(title = "Correlation Matrix", x = "", y = "") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```
Given the results in the correlation matrix, we see a semi-strong correlation between the mental health expenditure index and the health expenditure percentage. We have, therefore, decided not to include mental health expenditure index in any models, to avoid multicolinearity. 

# Modelling (AS)
```{r}
# Creating the first and most simple model
simple_model <- lm(Suicide_p100 ~ HExp_Pctage_Y, WHO_MHExp_and_Deaths)

summary(simple_model)
```
Health expenditure percentage seems to have an effect on the suicide rate, but there is still a lot of unexplained variance.



## Adding Random Effects (SBS)
```{r}
#install.packages("lmerTest")
#install.packages("MuMIn")
library(MuMIn)
library(lmerTest)

# Adding country as a random intercept
mixed_model <- lmer(Suicide_p100 ~ HExp_Pctage_Y+(1|Country_Name), WHO_MHExp_and_Deaths)
summary(mixed_model)

# Calculating the marginal and conditional R-sqaured
r_squared <- r.squaredGLMM(mixed_model)
print(r_squared)


```
Adding a random intercept for country has made the effect of health expenditure percentage more clear/significant, thus explaining 88% of the data.


```{r}
# Adding continent as a column
data <- WHO_MHExp_and_Deaths %>%
  count(continents)

# Counting the occurrences of each continent and calculating the percentage
data <- WHO_MHExp_and_Deaths %>%
  count(continents) %>%
  mutate(percentage = n / sum(n) * 100)

# Creating a bar plot to visualise the imbalance in the data
ggplot(data, aes(x = continents, y = percentage, fill = continents)) +
  geom_bar(stat = "identity", width = 0.7) +
  labs(title = "Distribution of Continents in the Dataset", 
       x = "Continent", 
       y = "Percentage (%)") +
  scale_fill_brewer(palette = "Set3") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5) +
  theme_minimal()

```
The bar plot shows that Europe is very prominent in the data, compared to the other countries, thus telling us that we should take the results with a grain of salt. 

```{r}
# recalculating the "Year" column to use in models
WHO_MHExp_and_Deaths$Year_since_start <- WHO_MHExp_and_Deaths$Year-2000

# Creating a plot with the Year as a predictor
ggplot(WHO_MHExp_and_Deaths, aes(x = Year_since_start, y = HExp_Pctage_Y)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", color = "blue", se = TRUE) + 
  labs(
    title = "Health Expenditure Percentage vs. Years Since Start",
    x = "Years Since Start",
    y = "Health Expenditure (% of GDP)"
  ) +
  theme_minimal() 
```
Visual inspection shows a slightly positive slope between the year and the health expenditure percentage.

## Adding more Predictors (LR)

```{r}
# Adding continent as an interaction term
continent_simple_model <- lm(Suicide_p100 ~ HExp_Pctage_Y+HExp_Pctage_Y:continents ,WHO_MHExp_and_Deaths)
summary(continent_simple_model)

# Calculating the marginal and conditional R-sqaured
r_squared <- r.squaredGLMM(continent_simple_model)
print(r_squared)

# Adding continent as both an interaction term and as a fixed predictor
continent_model <- lm(Suicide_p100 ~ HExp_Pctage_Y*continents ,WHO_MHExp_and_Deaths)
summary(continent_model)

# Calculating the marginal and conditional R-sqaured
r_squared <- r.squaredGLMM(continent_model)
print(r_squared)
```
The explainability of both models do not seem as strong as the models with countries as random intercepts.

# Model Selection (SBS)
```{r}
#using AIC to select the best and simplest model
AIC(simple_model, continent_model,continent_simple_model, mixed_model)
```
We choose the "mixed_model" as it has best explainability while having the lowest AIC score. 
